---
title: "Democracy & Gender in Leadership"
format:
  dashboard:
    orientation: columns
    theme: flatly
    embed-resources: true
    css: dashboard-styles.css
    logo: logo.png
    logo-alt: "Democracy and Gender in Leadership project logo."
nav-buttons:
  - icon: github
    href: https://github.com/japtekar89/Geopolitical-Analysis
    aria-label: GitHub
  - icon: linkedin
    href: https://www.linkedin.com/in/bryce-corbett-7640403a0/
    aria-label: LinkedIn
  - icon: instagram
    href: https://www.instagram.com/calpoly
    aria-label: Instagram
  - icon: envelope
    href: mailto:japtekar@calpoly.edu
    aria-label: Email
server: shiny
editor: visual
---

```{r}
#| context: setup
options(pillar.width = Inf)
library(tidyverse)
library(wesanderson)
library(gt)
library(broom)
library(sf)
library(leaflet)
library(shiny)
library(plotly)

democracy_data <- read.csv("Combined_Democracy_Data.csv")

blue_dark <- "#5b8db8"
blue_mid <- "#89afd4"
blue_light <- "#bdd4e7"
blue_pale <- "#ddeaf5"
pink_accent <- "#e8a0b0"
pink_pale <- "#f5d5dc"
black <- "#000000"
white <- "#FFFFFF"

pal <- wes_palette("Moonrise3", n = 5)

theme_set(theme_minimal(base_size = 13, base_family = "Georgia"))

democracy_data <- democracy_data |>
  mutate(president_age_at_accession = president_accesion_year - president_birthyear,
    monarch_age_at_accession   = monarch_accession_year  - monarch_birthyear,
    has_female_leader = ifelse(is_female_president | is_female_monarch, 1, 0),
    decade = case_when(
      year < 1960 ~ "1950s",
      year < 1970 ~ "1960s",
      year < 1980 ~ "1970s",
      year < 1990 ~ "1980s",
      year < 2000 ~ "1990s",
      year < 2010 ~ "2000s",
      TRUE  ~ "2010s"))

summarise_leadership_gender_gap <- function(group_var) {
  democracy_data |>
    filter(!is.na(.data[[group_var]])) |>
    group_by(across(all_of(group_var))) |>
    summarise(n = n(),
      prop_female_president = mean(ifelse(!is_monarchy, is_female_president, NA), na.rm = TRUE),
      prop_male_president = mean(ifelse(!is_monarchy, !is_female_president, NA), na.rm = TRUE),
      gap_president_prop = prop_male_president - prop_female_president,
      avg_accession_age_female_president = mean(
        president_age_at_accession[is_presidential & is_female_president], na.rm = TRUE),
      avg_accession_age_male_president = mean(
        president_age_at_accession[is_presidential & !is_female_president], na.rm = TRUE),
      gap_president_accession_age = avg_accession_age_male_president - avg_accession_age_female_president,
      prop_female_monarch = mean(ifelse(is_monarchy, is_female_monarch, NA), na.rm = TRUE),
      prop_male_monarch = mean(ifelse(is_monarchy, !is_female_monarch, NA), na.rm = TRUE),
      gap_monarch_prop  = prop_male_monarch - prop_female_monarch,
      avg_accession_age_female_monarch = mean(
        monarch_age_at_accession[is_monarchy & is_female_monarch], na.rm = TRUE),
      avg_accession_age_male_monarch = mean(
        monarch_age_at_accession[is_monarchy & !is_female_monarch], na.rm = TRUE),
      gap_monarch_accession_age = avg_accession_age_male_monarch - avg_accession_age_female_monarch,
      .groups = "drop")
}

plotMe <- summarise_leadership_gender_gap("year")

democracy_data <- democracy_data |> arrange(country, year)
democracy_data$to_democracy <- 0
democracy_data$from_democracy <- 0

for (i in 2:nrow(democracy_data)) {
  if (democracy_data$country[i] == democracy_data$country[i - 1]) {
    current  <- democracy_data$is_democracy[i]
    previous <- democracy_data$is_democracy[i - 1]
    if (!is.na(current) && !is.na(previous)) {
      if (current && !previous)  democracy_data$to_democracy[i]   <- 1
      if (!current && previous)  democracy_data$from_democracy[i] <- 1
    }
  }
}

summary_data <- democracy_data |>
  group_by(year) |>
  summarize(prop_democracy = mean(is_democracy, na.rm = TRUE),
    transitions_in  = sum(to_democracy, na.rm = TRUE),
    transitions_out = sum(from_democracy, na.rm = TRUE),
    .groups = "drop")

democracy_data <- democracy_data |>
  mutate(ten_yrs = floor(year / 10) * 10)

summary_ten_yrs <- democracy_data |>
  group_by(ten_yrs) |>
  summarize(prop_democracy = mean(is_democracy, na.rm = TRUE),
    transitions_in  = sum(to_democracy,   na.rm = TRUE),
    transitions_out = sum(from_democracy, na.rm = TRUE),
    .groups = "drop")

topn_countries <- function(start_year = 1950, end_year = 2020, num_countries = 10) {
  if (start_year < 1950 | end_year > 2020 | start_year > end_year)
    stop("Year range must be between 1950 and 2020")

  data <- democracy_data |>
    filter(year >= start_year & year <= end_year) |>
    arrange(country_name, year) |>
    group_by(country_name) |>
    mutate(prev_year = lag(is_democracy),
      to_democracy = ifelse(is_democracy == 1 & prev_year == 0, 1, 0),
      from_democracy = ifelse(is_democracy == 0 & prev_year == 1, 1, 0)) |>
    ungroup()

  data |>
    group_by(country_name) |>
    summarize(transitions_in  = sum(to_democracy,   na.rm = TRUE),
      transitions_out = sum(from_democracy, na.rm = TRUE),
      total_transitions = transitions_in + transitions_out,
      .groups = "drop") |>
    arrange(desc(total_transitions)) |>
    slice_head(n = num_countries)
}
```

# Introduction

## Column {width="50%"}

::: {.card title="About This Project"}
### Democracy & Gender in Leadership

This dashboard explores the relationship between democratic governance and gender representation in political leadership across the globe from **1950 to 2020**. 

**Key Questions We Investigate:**

- Do democratic countries have higher rates of women in parliament compared to non-democratic countries?
- How have gender gaps in presidential and monarchical leadership changed over time?
- What does the global map of female political representation look like — and how has it shifted?
- How do individual countries like Honduras illustrate broader global trends?

**Data Sources:**

Our primary dataset includes information on each country’s democratic status, executive leadership gender, parliamentary representation, and regime transitions per year from 1950-2020. Honduras-specific data comes from the **World Bank Open Data** portal, covering over 1,300 development indicators from 1960–2025.
:::

## Column {width="50%"}
::: {.card title="Global Democratic Trend" style="margin-top: 1rem;"}
```{r}
summary_data |>
  ggplot(aes(x = year, y = prop_democracy)) +
  geom_area(fill = blue_pale, alpha = 0.7) +
  geom_line(color = blue_dark, linewidth = 1.2) +
  scale_y_continuous(labels = scales::label_percent(), limits = c(0, 0.8)) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
  labs(x = NULL, y = "% of Countries That Are Democracies",
       title = "Rise of Democracy, 1950–2020") +
  theme(panel.grid.minor = element_blank(),
    axis.title = element_text(color = blue_dark),
    axis.text  = element_text(color = blue_dark),
    plot.title = element_text(color = blue_dark, size = 12))
```
:::

# Gender Gaps in Leadership and Representation

## Column {width="50%"}

:::: {.card title="Visualizations"}
::: panel-tabset
### Women in Parliament by Democratic Status

```{r}
democracy_data |>
  filter(!is.na(is_democracy), !is.na(prop_in_par)) |>
  mutate(is_democracy = factor(is_democracy, labels = c("Non-Democracy", "Democracy"))) |>
  ggplot(aes(x = is_democracy, y = prop_in_par)) +
  geom_boxplot(
    fill = c(blue_mid, blue_dark),
    color = black, outlier.color = black) +
  stat_summary(fun = mean, geom = "crossbar", color = black,
               width = 0.75, linetype = "dotted", linewidth = 0.3) +
  stat_summary(fun = mean, geom = "text",
               aes(label = paste0("Mean: ", round(after_stat(y), 1), "%")),
               vjust = -1, color = "white", size = 3.5) +
  stat_summary(fun = median, geom = "text",
               aes(label = paste0("Median: ", round(after_stat(y), 1), "%")),
               vjust = 1.5, color = "white", size = 3.5) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  labs(x = "Democracy Status", y = "% of Parliament Seats Held by Women") +
  theme(panel.grid.minor = element_blank(),
    axis.title = element_text(color = blue_dark),
    axis.text  = element_text(color = blue_dark))
```

### Average Accession Ages

No female presidents were recorded before 1980, so this plot begins there.

```{r}
plotMe |>
  filter(year >= 1980) |>
  ggplot(aes(x = year)) +
  geom_ribbon(aes(ymin = avg_accession_age_female_president,
                  ymax = avg_accession_age_male_president),
              fill = "#D3D3D3", alpha = 0.8) +
  geom_ribbon(aes(ymin = avg_accession_age_female_monarch,
                  ymax = avg_accession_age_male_monarch),
              fill = "#D3D3D3", alpha = 0.8) +
  geom_line(aes(y = avg_accession_age_female_president, color = "Female President"), linewidth = 1.4) +
  geom_line(aes(y = avg_accession_age_male_president, color = "Male President"), linewidth = 1.4) +
  geom_line(aes(y = avg_accession_age_female_monarch, color = "Female Monarch"), linewidth = 1.4) +
  geom_line(aes(y = avg_accession_age_male_monarch, color = "Male Monarch"), linewidth = 1.4) +
  scale_color_manual(
    name = "Leader Type",
    values = c("Female President" = pink_accent, "Male President" = blue_dark, "Female Monarch" = pink_accent, "Male Monarch"   = blue_mid),
    breaks = c("Female President", "Male President", "Female Monarch", "Male Monarch")) +
  scale_x_continuous(limits = c(1980, 2020)) +
  scale_y_continuous(limits = c(20, 70)) +
  labs(x = "Year", y = "Average Accession Age", color = "Leader Type") +
  theme(panel.grid.minor  = element_blank(),
    legend.position = "bottom",
    legend.direction  = "horizontal",
    legend.background = element_rect(fill = "white", color = blue_light),
    legend.text = element_text(size = 9, color = blue_dark),
    legend.title  = element_text(size = 9, color = blue_dark),
    axis.title = element_text(color = blue_dark),
    axis.text = element_text(color = blue_dark))
```
:::
::::

## Column {width="50%"}

:::::: {.card title="Tables"}
::::: panel-tabset
### Women in Parliament

```{r}
result  <- t.test(prop_in_par ~ is_democracy, data = democracy_data)
table_result <- tidy(result)

table_result |>
  select(estimate1, estimate2, statistic, p.value, conf.low, conf.high, parameter) |>
  gt() |>
  tab_header(title = md("**Table 1: Welch Two-Sample t-Test**"),
    subtitle = "Proportion of Parliament Seats Held by Women") |>
  cols_label(
    estimate1 = md("Mean<br>(Non-Dem.)"),
    estimate2 = md("Mean<br>(Dem.)"),
    statistic = md("*t*"),
    p.value = md("*p*-value"),
    conf.low  = "CI Low",
    conf.high = "CI High",
    parameter = md("*df*")) |>
  fmt_number(columns = c(estimate1, estimate2, statistic, conf.low, conf.high), decimals = 3) |>
  fmt_number(columns = parameter, decimals = 1) |>
  fmt_scientific(columns = p.value, decimals = 3) |>
  tab_style(style = list(cell_fill(color = pink_pale), cell_text(weight = "bold", color = blue_dark)),
    locations = cells_body(columns = p.value)) |>
  tab_style(style = cell_fill(color = blue_pale),
    locations = list(cells_column_spanners(), cells_column_labels(), cells_title())) |>
  tab_style(style = cell_text(color = blue_dark),
    locations = list(cells_column_labels(), cells_title(), cells_body())) |>
  tab_footnote(footnote= "Highlighted p-value is statistically significant (p < 0.05).",
    locations = cells_column_labels(columns = p.value)) |>
  tab_options(table.font.size = px(13),
    data_row.padding = px(5),
    table.background.color = "white",
    column_labels.background.color = blue_pale) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

::: {.card style="background-color: #f7fbff; border: 1px solid #bdd4e7; padding: 1rem; margin-top: 1rem;"}
The t-test (t = -4.591, p \< 0.001) confirms a statistically significant difference in women's parliamentary representation between democratic and non-democratic countries. Democratic countries average **17.7%** versus **16.0%** for non-democracies, with a confidence interval that excludes zero.
:::

### Accession Ages Over Time

```{r}
summarise_leadership_gender_gap("is_democracy") |>
  mutate(is_democracy = ifelse(is_democracy, "Democracy", "Non-Democracy")) |>
  select(is_democracy, n,
    avg_accession_age_female_president,
    avg_accession_age_male_president,
    gap_president_accession_age,
    avg_accession_age_female_monarch,
    avg_accession_age_male_monarch,
    gap_monarch_accession_age) |>
  gt() |>
  tab_header(title = md("**Table 2: Gender Gaps in Accession Age**"),
    subtitle = "Average Age When Leaders First Take Office") |>
  tab_spanner(label = "Presidents",
    columns = c(avg_accession_age_female_president, avg_accession_age_male_president,
                gap_president_accession_age)) |>
  tab_spanner(label = "Monarchs",
    columns = c(avg_accession_age_female_monarch, avg_accession_age_male_monarch,
                gap_monarch_accession_age)) |>
  cols_label(is_democracy = "Status",
    n = "N",
    avg_accession_age_female_president = "Female",
    avg_accession_age_male_president = "Male",
    gap_president_accession_age = "Gap (M-F)",
    avg_accession_age_female_monarch = "Female",
    avg_accession_age_male_monarch = "Male",
    gap_monarch_accession_age = "Gap (M-F)") |>
  fmt_number(columns = c(avg_accession_age_female_president, avg_accession_age_male_president, gap_president_accession_age, avg_accession_age_female_monarch, avg_accession_age_male_monarch, gap_monarch_accession_age),
    decimals = 1) |>
  fmt_integer(columns = n) |>
  tab_style(style = cell_fill(color = blue_pale),
    locations = list(cells_column_spanners(), cells_column_labels(), cells_title())) |>
  tab_style(style = cell_text(color = blue_dark),
    locations = list(cells_column_labels(), cells_title(), cells_body())) |>
  tab_style(style = cell_fill(color = pink_pale),
    locations = cells_body(
      columns = c(gap_president_accession_age, gap_monarch_accession_age),
      rows= gap_president_accession_age > 0)) |>
  tab_style(style = cell_fill(color = blue_light),
    locations = cells_body(
      columns = c(gap_president_accession_age, gap_monarch_accession_age),
      rows = gap_president_accession_age < 0)) |>
  tab_footnote(footnote = "Blue = males higher; pink = females higher.",
    locations = cells_column_labels(columns = c(gap_president_accession_age, gap_monarch_accession_age))) |>
  tab_options(table.font.size = px(13),
    data_row.padding = px(5),
    table.background.color = "white",
    column_labels.background.color = blue_pale) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

::: {.card style="background-color: #f7fbff; border: 1px solid #bdd4e7; padding: 1rem; margin-top: 1rem;"}
In non-democratic countries, female presidents take office nearly **14 years older** than male presidents on average. In democracies, that gap shrinks to under 3 years. For monarchs, the pattern reverses — female monarchs tend to ascend younger, and the gap has been growing since 2000.
:::

### Gender Gaps by Decade

```{r}
pal3 <- wes_palette("Moonrise3", n = 4)

summarise_leadership_gender_gap("decade") |>
  select(decade, n, prop_female_president, prop_male_president, gap_president_prop) |>
  gt() |>
  tab_header(title = md("**Table 3: Gender Gaps by Decade**")) |>
  cols_label(decade = "Decade", n = "Sample Size",
    prop_female_president = "Prop. Female President",
    prop_male_president = "Prop. Male President",
    gap_president_prop = "Gap (M - F)") |>
  fmt_percent(columns = c(prop_female_president, prop_male_president, gap_president_prop),
    decimals = 1) |>
  fmt_integer(columns = n) |>
  tab_style(style = cell_fill(color = blue_pale),
    locations = list(cells_column_labels(), cells_title())) |>
  tab_style(style = cell_text(color = blue_dark),
    locations = list(cells_column_labels(), cells_title(), cells_body())) |>
  tab_style(style = cell_fill(color = paste0(pal3[2], "99")),
    locations = cells_body(columns = gap_president_prop, rows = gap_president_prop > 0)) |>
  tab_style(style = cell_fill(color = paste0(pal3[3], "99")),
    locations = cells_body(columns = gap_president_prop, rows = gap_president_prop < 0)) |>
  tab_footnote(footnote= "Positive gap means more male than female presidents. Gap narrows across decades.",
    locations = cells_column_labels(columns = gap_president_prop)) |>
  tab_options(table.font.size = px(13),
    data_row.padding = px(5),
    table.background.color = "white",
    column_labels.background.color = blue_pale) |>
  opt_table_font(font = google_font("Source Serif 4"))
```
:::::
::::::

# Interactive Visualizations {orientation="columns"}

```{r}
country_summaries <- democracy_data |>
  arrange(country, year, country_name) |>
  group_by(country) |>
  summarize(country_name = first(country_name),
    avg_prop_in_par = mean(prop_in_par, na.rm = TRUE),
    pct_years_democratic = mean(is_democracy, na.rm = TRUE),
    total_transitions = sum(to_democracy + from_democracy, na.rm = TRUE),
    early_par = mean(prop_in_par[year <= 2005], na.rm = TRUE),
    late_par  = mean(prop_in_par[year >= 2010], na.rm = TRUE),
    change_in_par = late_par - early_par,
 .groups = "drop") |>
  filter(
    !is.nan(avg_prop_in_par),
    !is.nan(early_par),
    !is.nan(late_par),
    !is.nan(change_in_par),
    !is.na(pct_years_democratic))

global_avg_par <- mean(country_summaries$avg_prop_in_par, na.rm = TRUE)
global_avg_dem  <- mean(country_summaries$pct_years_democratic, na.rm = TRUE)

representative <- country_summaries |>
  mutate(distance_from_avg = abs(avg_prop_in_par - global_avg_par) +
                        abs(pct_years_democratic - global_avg_dem)) |>
  arrange(distance_from_avg) |>
  slice_head(n = 5)

honduras_clean <- read_csv("API_HND_DS2_en_csv_v2_19376.csv", skip = 4) |>
  select(-last_col()) |>
  filter(!is.na(`Indicator Name`)) |>
  pivot_longer(
    cols = `1960`:`2025`,
    names_to = "year",
    values_to = "value") |>
  mutate(year= as.integer(year),
    value = as.numeric(value)) |>
  filter(!is.na(value)) |>
  rename(indicator_name = `Indicator Name`) |>
  select(indicator_name, year, value) |>
  pivot_wider(names_from = indicator_name,
    values_from = value) |>
  arrange(year)

honduras_plot <- honduras_clean |>
  select(year,
    women_parliament = "Proportion of seats held by women in national parliaments (%)",
    labor_force_female = "Labor force participation rate, female (% of female population ages 15+) (modeled ILO estimate)") |>
  filter(!is.na(women_parliament) | !is.na(labor_force_female))
```

## Column {width="65%"}

::: card
```{r}
plot_ly(honduras_plot, x = ~year) |>
  add_lines(y = ~women_parliament,
    name = "Women in Parliament (%)",
    line = list(color = "#899DA4", width = 2),
    yaxis = "y1") |>
  add_lines(y = ~labor_force_female,
    name = "Female Labor Force Participation (%)",
    line = list(color = "#C93312", width = 2),
    yaxis = "y2") |>
  layout(title = list(
      text = "Women in Parliament and Female Labor Force Participation in Honduras",
      x = 0.5,
      font = list(size = 13)),
    xaxis = list(title = "Year"),
    yaxis = list(
      title = "Women in Parliament (%)",
      titlefont = list(color = "#899DA4"),
      tickfont = list(color = "#899DA4"),
      range = list(0, 35)),
    yaxis2 = list(title = "Female Labor Force Participation (%)",
      titlefont = list(color = "#C93312"),
      tickfont = list(color = "#C93312"),
      overlaying = "y",
      side = "right",
      range = list(20, 60)),
    legend  = list(x = 0.05, y = 0.95),
    hovermode = "x unified",
    margin = list(r = 60, t = 55))
```
:::

::: {.card title="Map of Female Political Representation"}
```{r}
#| context: setup

world_sf <- map_data("world") %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326) %>%
  group_by(region, group) %>%
  summarise(geometry = st_combine(geometry), .groups = "drop") %>%
  st_cast("POLYGON") %>%
  group_by(region) %>%
  summarise(geometry = st_combine(geometry), .groups = "drop") %>%
  st_cast("MULTIPOLYGON") %>%
  mutate(country_name = recode(
    region,
    "USA" = "United States",
    "UK" = "United Kingdom",
    "Democratic Republic of the Congo" = "Congo, Dem. Rep.",
    "Republic of Congo" = "Congo, Republic of",
    "Ivory Coast" = "Côte d`Ivoire",
    "Gambia" = "Gambia, The",
    "North Korea" = "Korea, People's Republic",
    "South Korea" = "Korea, Republic of",
    "Slovakia" = "Slovak Republic",
    "Saint Kitts and Nevis" = "St. Kitts & Nevis",
    "Saint Lucia" = "St. Lucia",
    "Saint Vincent and the Grenadines" = "St.Vincent & Grenadines",
    "Trinidad and Tobago" = "Trinidad &Tobago",
    "Micronesia" = "Micronesia, Fed. Sts.",
    "Turks and Caicos Islands" = "Turks and Caicos"
  ))

marker_coords <- map_data("world") %>%
  group_by(region) %>%
  summarize(
    long = mean(range(long, na.rm = TRUE)),
    lat = mean(range(lat, na.rm = TRUE)),
    .groups = "drop"
  )

pinks <- colorNumeric("RdPu", domain = c(0, 100), na.color = "#cccccc")

marker_pal <- colorFactor(
  palette = c(blue_dark, blue_pale, black),
  levels = c("President", "Monarch", "Both"))
```

```{r}
sliderInput(
  "year",
  label    = "Select Year:",
  min      = 1950,
  max      = 2020,
  value    = 2020,
  step     = 1,
  sep      = "",
  width    = "100%",
  animate  = animationOptions(interval = 300, loop = FALSE)
)

leafletOutput("map", height = 530)

tags$div(
  style = "color: #5b8db8; font-size: 0.9rem; font-weight: 600; text-align: center; margin-top: 0.5rem;",
  "Global distribution of women in parliament and countries with female presidents or monarchs, 1950–2020")
```
:::

## Column {width="50%"}

::: {.card title="Gender Gaps by Country and Decade"}
```{r}
selectInput("country",
  "Select a country:",
  choices = sort(unique(democracy_data$country_name)),
  selected = sort(unique(democracy_data$country_name))[1])

gt_output("country_gap_table")
```
:::

```{r}
#| context: server

output$map <- renderLeaflet({
  world_year <- world_sf %>%
    left_join(democracy_data %>% filter(year == input$year),
      by = "country_name")

  female_leaders <- world_year %>%
    filter(is_female_president | is_female_monarch) %>%
    mutate(leader_type = case_when(
        is_female_president & is_female_monarch ~ "Both",
        is_female_president ~ "President",
        is_female_monarch ~ "Monarch")) %>%
    st_drop_geometry() %>%
    left_join(marker_coords, by = "region") %>%
    filter(!is.na(long), !is.na(lat))

  leaflet(world_year) %>%
    addProviderTiles(providers$CartoDB.Positron,  
options = providerTileOptions(noWrap = TRUE)) %>%
    setView(lng = 0, lat = 20, zoom = 2) %>%
    addPolygons(fillColor = ~pinks(prop_in_par),
      fillOpacity = 1,
      color = "#555555",
      weight = 0.5,
      label = ~paste0(
        region,
        " | Women in parliament: ", round(prop_in_par, 1),
        "% | Democracy: ", ifelse(is_democracy, "Yes", "No"))) %>%
    addCircleMarkers(data = female_leaders,
      lng = ~long,
      lat = ~lat,
      radius = 6,
      stroke = TRUE,
      weight = 1,
      color = "#222222",
      fillColor = ~marker_pal(leader_type),
      fillOpacity = 0.9,
      label = ~paste(region, "-", leader_type)) %>%
    addLegend(
      pal = pinks,
      values = c(0, 100),
      title = NULL,
      position = "bottomright",
      labFormat = labelFormat(suffix = "%")) %>%
    addLegend(position = "topright",
      pal = marker_pal,
      values = c("President", "Monarch", "Both"),
      title = NULL)
})

output$country_gap_table <- render_gt({
  democracy_data %>%
    filter(country_name == input$country) %>%
    group_by(decade) %>%
    summarise(
      prop_female_president = mean(ifelse(!is_monarchy, is_female_president, NA), na.rm = TRUE),
      prop_male_president = mean(ifelse(!is_monarchy, !is_female_president, NA), na.rm = TRUE),
      gap_president_prop = prop_male_president - prop_female_president,
      prop_female_monarch = mean(ifelse(is_monarchy, is_female_monarch, NA), na.rm = TRUE),
      prop_male_monarch = mean(ifelse(is_monarchy, !is_female_monarch, NA), na.rm = TRUE),
      gap_monarch_prop = prop_male_monarch - prop_female_monarch,
      .groups = "drop") %>%
    gt() %>%
    tab_header(
      title = md(paste0("**Gender Gaps in Executive Leadership: ", input$country, "**")),
      subtitle = "Average Percentages of Female and Male Leaders by Decade") %>%
    tab_spanner(
      label = "Presidents",
      columns = c(prop_female_president, prop_male_president, gap_president_prop)) %>%
    tab_spanner(
      label = "Monarchs",
      columns = c(prop_female_monarch, prop_male_monarch, gap_monarch_prop)) %>%
    cols_label(
      decade = "Decade",
      prop_female_president = "Female",
      prop_male_president = "Male",
      gap_president_prop = "Gap",
      prop_female_monarch = "Female",
      prop_male_monarch = "Male",
      gap_monarch_prop = "Gap") %>%
    fmt_percent(
      columns = c(
        prop_female_president,
        prop_male_president,
        gap_president_prop,
        prop_female_monarch,
        prop_male_monarch,
        gap_monarch_prop),
      decimals = 1) %>%
    cols_align(align = "center", columns = everything()) %>%
    tab_style(
      style = cell_fill(color = blue_pale),
      locations = list(cells_column_spanners(), cells_column_labels(), cells_title())) %>%
    tab_style(
      style = cell_text(color = blue_dark),
      locations = list(cells_column_labels(), cells_title(), cells_body())) %>%
    tab_source_note(
  source_note = "NaN values indicate that data were unavailable for that decade or leadership category.") |>
    opt_table_font(font = "Source Serif 4")
})
```

# Conclusion

## Column {width="40%"}

:::: {.card title="Honduras Trends"}
::: panel-tabset
### Honduras as a Representative Case

Honduras had 6 transitions in and out of democracy. The difference from the average proportion of women in parliament from before 2005 and after 2020 was 0.5212923, which is pretty close to the average as well. So, it both represents the global trend for women in parliament pretty well, but acts as an outlier for the number of transitions in and out of democracy.

### Women in Parliament Trends

The female population in Honduras remains relatively constant over time. However, the percentage of women in parliament fluctuates throughout the years. In 1997, only about 9% of seats in parliament were occupied by women, and in 2002, this percentage dropped to around 5.5%. 2005 saw a significant increase of 23% in the percentage of women in parliament. This percentage has varied since then, however the percentage of women in parliament seems to be on the rise, with almost 30% of seats being held by women in 2025.

### Global Trends

South America has a large concentration of female leaders. There are also a few European, African, and Asian countries with either female monarchs or presidents. Generally, it seems that smaller territories and countries, such as the Cayman Islands and Nepal, have female leaders more than larger countries. The map also reveals that countries with female leaders don't necessarily have a large proportion of women in parliament. For example, Liberia, South Korea, and Greece are some countries who once had female leaders, but have low proportions of women in parliament.
:::
::::

## Column {width="60%"}

::: {.card title="Key Findings"}
```{r}
# Visual summary of key findings as a styled HTML block
findings <- tibble(
  heading = c(
    "Democracy & Women in Parliament",
    "Presidential Gender Gap Narrows",
    "Monarchies: A Different Pattern",
    "Democratization Is Not Linear",
    "Progress Is Real But Uneven"
  ),
  detail  = c(
    "Democratic countries averaged 17.7% women in parliament vs. 16.0% in non-democracies — a statistically significant difference (p < 0.001), though the gap is modest.",
    "In non-democracies, female presidents take office nearly 14 years later than male counterparts. In democracies, this accession-age gap shrinks to under 3 years.",
    "Female monarchs tend to come to power younger than male monarchs, and this gap has widened since 2000, opposite from the presidential pattern.",
    "The share of democratic countries grew substantially from 1950 to 2020, especially in the late 1980s and early 1990s. However, countries continued moving both into and out of democracy throughout the years.",
    "Female presidents became more common each decade, yet male presidents still made up the vast majority of leaders across all decades studied."
  )
)

findings |>
  gt() |>
  cols_label(heading = "Finding", detail = "Detail") |>
  tab_header(title = md("**Summary of Key Findings**")) |>
  tab_style(style = cell_fill(color = blue_pale),
    locations = list(cells_column_labels(), cells_title())) |>
  tab_style(style = cell_text(color = blue_dark, weight = "bold"),
    locations = cells_body(columns = heading)) |>
  tab_style(style = cell_text(color = blue_dark),
    locations = list(cells_column_labels(), cells_title(), cells_body(columns = detail))) |>
  tab_style(style = cell_fill(color = blue_pale),
    locations = cells_body(rows = c(1, 3, 5))) |>
  tab_options(table.font.size = px(13),
    data_row.padding = px(10),
    table.background.color = "white",
    column_labels.background.color = blue_pale,
    table.border.top.color = blue_light,
    table.border.bottom.color = blue_light) |>
  opt_table_font(font = google_font("Source Serif 4"))
```
:::

::: {.card style="background-color: #f7fbff; border: 1px solid #bdd4e7; padding: 1rem; margin-top: 1rem;"}
Overall, we found that democracy is related to gender representation and political stability, but not in simple ways. Democratic countries had slightly higher average proportions of women in parliament than non-democratic countries, and the t-test showed that this difference was statistically significant. However, the actual difference in means was fairly small, suggesting that democracy alone does not guarantee equal representation. Female presidents became more common over time, but male presidents still made up the vast majority of presidents across every decade. Accession-age gaps were smaller and more stable among presidents, while monarchs showed larger differences by gender.

Our analysis of democratic transitions reveals that global democratization has been unpredictable. The percentage of democratic countries increased substantially from 1950 to 2020, especially around the late 1980s and early 1990s, but this growth was not steady. Countries continued to move both into and out of democracy over time, showing that democratization includes both progress and reversal. The country-level transition tables suggest that even a small number of regime changes can reflect meaningful political instability.

Taken together, these findings show that becoming a democratic country can be associated with greater gender representation and broader political change, but maintaining democratic rule and achieving gender equality remain ongoing challenges. Democracy may create more opportunities for representation, but the persistence of gender gaps and repeated regime transitions show that political progress is not automatic or permanent.
:::


