---
title: "Project 1 Checkpoint 2"
format: html
---
```{r}
library(tidyverse)
```

```{r}
democracy_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-11-05/democracy_data.csv')
```

## Context
This is a long data frame with 14,768 observations and 43 variables. 
Each row contains political information for one country during a particular year.
The variables of the data frame provide information on whether a country is democratic, what type of regime the country has, the structure of the political system, along with information on elections, leaders, suffrage, power changes, and colonial history. 
There are variables about whether the leader is a president or a monarch, when their leadership term began, their gender, and their birth year. 
Three variables provide information on timing of regime or leadership changes. 
Lastly, three variables provide information on colonial status and relationships of a country.

## What Cleaning Has Been Done?
First, variables were renamed to all lowercase names with clean_names(). Then, many variable names were renamed to clearer and more descriptive names (for example, dd_regime became regime_category_index). Next, "?" values in the election_month column were converted to NA with na_if(), before the column was split into separate month and year variables with separate_wider_regex(). A new variable, electoral_category, is created using case_match() from electoral_category_index which contained values of 0-3 that measure the level of electoral competition; a 0 means no elections, a 1 means single-party elections, a 2 means non-democratic multi-party elections, and a 3 means democratic elections. Also, str_squish() was used to clean up whitespaces in the election_month variable. Lastly, variables were converted to appropriate data types using across(); year/count columns became numeric and indicator variables (starting with _is or _has) became logical.


**Question 1:** Among countries in the year 2000, how does the proportion of full suffrage differ across electoral systems (monarchy vs. presidential)?

* No new column with custom function
```{r}
full_sample_2000s <- democracy_data %>%
  filter(year == 2000) %>%
  distinct(country_name, .keep_all = TRUE)

monarchy_2000s <- democracy_data %>%
  filter(is_monarchy, year == 2000) %>%
  distinct(country_name, .keep_all = TRUE)

presidential_2000s <- democracy_data %>%
  filter(is_presidential, year == 2000) %>%
  distinct(country_name, .keep_all = TRUE)

prop_full_suffrage <- function(df) {
  mean(df$has_full_suffrage, na.rm = TRUE)
}

overall_prop <- prop_full_suffrage(full_sample_2000s)
monarchy_prop <- prop_full_suffrage(monarchy_2000s)
presidential_prop <- prop_full_suffrage(presidential_2000s)

data.frame(
  system = c("Overall", "Monarchy", "Presidential"),
  proportion_full_suffrage = c(overall_prop, monarchy_prop, presidential_prop)
)
```
**Conclusion**

Looking at the summary statistics for the year 2000, presidential countries are more likely to have full suffrage compared to monarchies. This result makes sense because presidents are typically elected by the public, while monarchs usually inherit their positions through hereditary systems.

The data shows that approximately **95.6%** of all countries in the sample had full suffrage in 2000, compared to **82.0%** of monarchies and **100%** of presidential systems. This indicates that presidential systems were almost universally associated with full suffrage at that time.

However, this is a correlation and not causation, meaning that the type of electoral system does not necessarily cause differences in suffrage.

**Question 2:**

* Modified Columns and a Custom Function

* We created four new variables.

- president_age_at_accession: a president’s age at their accession, calculated by subtracting the president’s birth year from their accession year

- monarch_age_at_accession: a monarch’s age at their accession, calculated by subtracting the monarch’s birth year from their accession year

- has_female_leader: an indicator variable indicating whether or not a country had a female president or monarch for that year, a 1 is “yes” and a 0 is “no”

- decade: decade category, split up the years evenly using nested ifelse() statements

* Afterwards, we wrote a custom function summarise_leadership_gender_gap() that uses an input grouping variable to group the data before calculating proportions of female and male monarchs and presidents, their average accession ages, and the difference in those proportions and ages.

```{r}
# creating our new variables
democracy_data <- democracy_data |>
  mutate(
    president_age_at_accession = president_accesion_year - president_birthyear,
    monarch_age_at_accession = monarch_accession_year - monarch_birthyear,
    has_female_leader = ifelse(is_female_president | is_female_monarch, 1, 0),
    decade = ifelse(year < 1960, "1950-1959",
       ifelse(year < 1970, "1960-1969",
       ifelse(year < 1980, "1970-1979",
       ifelse(year < 1990, "1980-1989",
       ifelse(year < 2000, "1990-1999",
       ifelse(year < 2010, "2000-2009",
                           "2010-2020")))))))

# writing our custom function
summarise_leadership_gender_gap <- function(group_var) {
  democracy_data |> 
    filter(!is.na(democracy_data[[group_var]])) |>
    # grouping data by specified grouping variable
    group_by(across(all_of(group_var))) |>
    summarise(n = n(),
      # calculating proportions of female and male presidents
      prop_female_president = mean(
        ifelse(!is_monarchy, is_female_president, NA), na.rm = TRUE),
      prop_male_president = mean(
        ifelse(!is_monarchy, !is_female_president, NA), na.rm = TRUE),
      # calculating difference in proportions of male presidents and female presidents (male - female)
      gap_president_prop = prop_male_president - prop_female_president,
      # calculating the average accession ages for female and male presidents
      avg_accession_age_female_president = mean(
        president_age_at_accession[is_presidential & is_female_president], na.rm = TRUE),
      avg_accession_age_male_president = mean(
        president_age_at_accession[is_presidential & !is_female_president], na.rm = TRUE),
      # calculating difference in average accession ages of male and female presidents (male - female)
      gap_president_accession_age = avg_accession_age_male_president - avg_accession_age_female_president,
      
      # calculating proportions of female and male monarchs
      prop_female_monarch = mean(
        ifelse(is_monarchy, is_female_monarch, NA), na.rm = TRUE),
      prop_male_monarch = mean(
        ifelse(is_monarchy, !is_female_monarch, NA), na.rm = TRUE),
      # calculating difference in proportions of male monarchs and female monarchs (male - female)
      gap_monarch_prop = prop_male_monarch - prop_female_monarch,
      # calculating the average accession ages for female and male monarchs
      avg_accession_age_female_monarch = mean(
        monarch_age_at_accession[is_monarchy & is_female_monarch], na.rm = TRUE),
      avg_accession_age_male_monarch = mean(
        monarch_age_at_accession[is_monarchy & !is_female_monarch], na.rm = TRUE),
      # calculating difference in average accession ages of male and female monarchs (male - female)
      gap_monarch_accession_age =
        avg_accession_age_male_monarch - avg_accession_age_female_monarch,
      .groups = "drop")
}
```

We can add any variable from the dataset as a grouping variable. For example, we could input is_democracy or has_free_and_fair_election instead of era to see the different values for democratic vs non-democratic countries or countries with and without free and fair elections.

```{r}
summarise_leadership_gender_gap("is_democracy")
summarise_leadership_gender_gap("has_free_and_fair_election")
```

In particular, however, we want to see those different values by decade.

```{r}
summarise_leadership_gender_gap("decade")
```

**Conclusion**

Over time, the proportion of female presidents has increased to 6.96% of countries having a female president by 2020. The gender gap for presidents has decreased by 13.7 percentage points for female and male presidents. The average accession ages for female and male presidents are similar across all years, typically around 50-55 years old. The largest difference in accession ages occurred in 1970-1979, when the average accession age of female presidents was 43, and the average accession age of male presidents was 48.6.

The proportion of female monarchs was increasing until the 2000-2009 decade, where it slightly dropped from 35.6% to 32.6%. After that, it dropped to 25.3% for the 2010-2020 decade. This is the lowest proportion of female monarchs across all decades; thus the gender gap for monarchs increased from 41.7% in 1950-1959 to 49.5% in 2010-2020. Average accession ages for monarchs are much younger than that of presidents. The typical accession age for a monarch is around 30 years. Female monarchs seem to be slightly younger than male monarchs at accession. The most dramatic difference in accession ages is seen during 2010-2020, where the average male monarch accession age was 44.1, and the average female monarch accession age was 28.6. 

**Question 3:** How does the proportion of countries with full suffrage differ between monarchies and presidencies over each decade (1950s-2020s)?

* Iteration with no new columns

```{r}
decade_range <- seq(from = 1950, to = 2020, by = 10)

suff_results <- data.frame(decade = integer(),
                           gov_type = character(),
                           prop_full_suffrage = numeric())
for(i in decade_range){
  start_year <- i
  end_year <- i + 9
  
  dd_decade <- democracy_data |>
    filter(year >= start_year & year <= end_year)
  
for(j in c("Presidential democracy", "Royal dictatorship")){
  dd_subset <- dd_decade |>
    filter(regime_category == j) |>
    distinct(country_name, .keep_all = TRUE)
  
  if(nrow(dd_subset) > 0){
    prop <- sum(dd_subset$has_full_suffrage == TRUE, na.rm = TRUE)/nrow(dd_subset)
  }
  else{
    prop <- NA
  }
  suff_results <- rbind(suff_results, data.frame(
  decade = i,
  gov_type = j,
  prop_full_suffrage = prop
  ))
  }
}
suff_results
```

**Conclusion**

The results above indicate clear differences between countries under presidential democracies and countries under royal dictatorships. Beginning in the 1950s, 50% of countries under presidential democracies had full suffrage, while 0% of countries under royal dictatorships had full suffrage. The proportion of presidential democracies with full suffrage quickly increased. The proportion jumped to about 94% in the 1970s, and has been at/near 100% since the 1980s. However, the opposite is true for countries under royal dictatorships. From the 1960’s to the 2000s, only about 20-36% of countries had full suffrage. This proportion fluctuated throughout the decades. However, this proportion increased to about 75% in the 2010s. Ultimately, presidential democracies were quick to grant full suffrage, while royal dictatorships fall behind and are slower to grant full suffrage. 

**Question 4:** How do democratic transitions within countries contribute to changes in the global proportion of democratic regimes over time?

* modified) columns with iteration

* New column: to_democracy, from_democracy, ten_yrs

* Required Iteration: going through the data and changing the value of the columns to_democracy and from_democracy

```{r}
#arranging by country and year
democracy_data <- democracy_data |> 
  arrange(country_code, year)

#creating new columns to the data, and assigning them to 0 to start with. this is an indicator variable, with 0 being no transition, and 1 meaning there is a transition. for example, if for a country 2000 is a 0 for the to_democracy variable and 2001 there is a 1, there was a transition to democracy in 2001. 
democracy_data$to_democracy <- 0
democracy_data$from_democracy <- 0

#to iterate through rows to find out which ones have transitions. starting at 2 because we are comparing to the previous row (2-1 = 1)
for (i in 2:nrow(democracy_data)) {
  #checking if the two countries that are next to each other in the data are the same 
  if (democracy_data$country_code[i] == democracy_data$country_code[i-1]) { 
    #assigning them to current and previous labels
    current <- democracy_data$is_democracy[i]
    previous <- democracy_data$is_democracy[i - 1]
    #making sure there are no na values for the is_democracy variable
    if (!is.na(current) && !is.na(previous)) {
      #if this year there is democracy and the previous year there wasn't, it is marked as a transition to democracy
      if (current && !previous) {
        democracy_data$to_democracy[i] <- 1
      }
      #opposite for this one 
    if (!current && previous) {
        democracy_data$from_democracy[i] <- 1
      }
    }
  }
}

#first grouping by year, and then summarizing the proportion of democracy, and transitions in and out of democracy
summary_data <- democracy_data |>
  group_by(year) |>
  summarize(
    prop_democracy = mean(is_democracy, na.rm = TRUE),
    transitions_in = sum(to_democracy, na.rm = TRUE),
    transitions_out = sum(from_democracy, na.rm = TRUE)
  )
summary_data

#after viewing this data, it would be helpful to analyze these trends using something like a plot, but looking at it as a table is overwhelming. To counteract this, we can condense the years to 5 or 10 years at the time to get a better look at overall trends. 
#we have to create a new column to group the years
democracy_data <- democracy_data |>
  mutate(ten_yrs = floor(year / 10) * 10)
#used the floor function here, that rounds a number down to the greatest integer less than or equal to it, here by 10s.
summary_ten_yrs <- democracy_data |>
  group_by(ten_yrs) |>
  summarize(
    prop_democracy = mean(is_democracy, na.rm = TRUE),
    transitions_in = sum(to_democracy, na.rm = TRUE),
    transitions_out = sum(from_democracy, na.rm = TRUE),
    .groups = "drop"
  )
summary_ten_yrs

#we also want to look at the correlation between the to and from democracy;s and the proportion of countries that are democratic. 
cor_matrix <- summary_data |>
  select(prop_democracy, transitions_in, transitions_out) %>%
  cor(use = "complete.obs")

cor_matrix

```

**Conclusion**
Looking at this table, it is clear that democracy has been overwhelmingly on the rise. However, there have been the most transitions into democracy in the 1990s, and the most transitions out in the 1960s, which doesn't necessarily line up with the transition numbers. So, I created a correlation matrix to look at the relationship between the proportion of democratic countries and the flow of countries changing to be either democratic or not. They all had particularly weak correlations, and this means that there isn't necessarily a relationship between the overall amounts of democratic countries and the state of transitions that there are in the world. This means that the transitions to and from democracy often just offset each other. 




