---
title: "The Democracy Gap"
author: "Julia Aptekar, Meghana Kolar, Sofia Jachno, & Bryce Corbett"
embedded-resources: true
code-fold: true
format: html
theme: lux
---
```{r}
#| warning: false
#install.packages("tidyverse")
#install.packages("wesanderson")
#install.packages("gt")
#install.packages("broom")
options(pillar.width = Inf)
library(tidyverse)
library(wesanderson)
library(gt)
library(broom)
democracy_data <- read.csv("Combined_Democracy_Data.csv")
pal <- wes_palette("Moonrise3", n = 5)
```

## Explore the data

Our study utilized two datasets. The first was compiled by Christian Bjørnskov and Martin Rode, two economists who study political institutions and how governments change over time. Their goal was to create a resource for researchers studying political transitions, an area they felt had not been adequately covered by existing data. The dataset collects information on 192 sovereign countries and 16 self-governing territories annually from 1950 to 2020 on each country's regime type, democratic status, elections, leadership, suffrage, transfers of power, and colonial history. In total, it includes 14,768 country-year observations across 43 variables.

To build this dataset, the authors expanded on an existing collection of political data by recording key features of each country's government, such as whether elections took place, whether those elections were considered free and fair, and whether leadership changed hands peacefully. Records of coups were verified across newspaper archives, encyclopedias, and other databases to ensure accuracy. For our analysis, this dataset is especially useful because it gives us a reliable way to classify each country as democratic or not for any given year, while also providing the broader political context needed to understand how and why governments change over time.

The second dataset comes from the World Bank, which tracks global progress on the UN's 17 Sustainable Development Goals, a set of goals aimed at making the world more equitable and sustainable across areas like poverty, education, and gender equality. The 5th goal focuses specifically on gender equality, and one of its key indicators is the percentage of parliamentary seats held by women in each country. This figure is compiled by the World Bank using data from the Inter-Parliamentary Union, an organization that collects official seat counts directly from legislative bodies around the world. It is available annually for most countries from 1997 to 2020, with coverage becoming more consistent in the later years of that range.

**Read more about the data here:**

[Democracy Data](https://www.christianbjoernskov.com/bjoernskovrodedata/)

[World Bank Data](https://databank.worldbank.org/source/sustainable-development-goals-(sdgs)#)

## What We Want To Know

1. How does the democratic context of a country shape gender gaps in parliamentary representation and executive leadership over time?
2. How do democratic transitions within countries contribute to changes in the global proportion of democratic regimes over time?

## Exploring Gender Gaps in Leadership and Representation

```{r}
#| warning: false
democracy_data <- democracy_data |>
  mutate(president_age_at_accession = president_accesion_year - president_birthyear,
    monarch_age_at_accession = monarch_accession_year - monarch_birthyear,
    has_female_leader = ifelse(is_female_president | is_female_monarch, 1, 0),
    decade = ifelse(year < 1960, "1950-1959",
       ifelse(year < 1970, "1960-1969",
       ifelse(year < 1980, "1970-1979",
       ifelse(year < 1990, "1980-1989",
       ifelse(year < 2000, "1990-1999",
       ifelse(year < 2010, "2000-2009",
                           "2010-2020")))))))

summarise_leadership_gender_gap <- function(group_var) {
  democracy_data |> 
    filter(!is.na(democracy_data[[group_var]])) |>
    # grouping data by specified grouping variable
    group_by(across(all_of(group_var))) |>
    summarise(n = n(),
              # calculating proportions of female and male presidents
              prop_female_president = mean(
                ifelse(!is_monarchy, is_female_president, NA), na.rm = TRUE),
              prop_male_president = mean(
                ifelse(!is_monarchy, !is_female_president, NA), na.rm = TRUE),
              # calculating difference in proportions of male presidents and female presidents (male - female)
              gap_president_prop = prop_male_president - prop_female_president,
              # calculating the average accession ages for female and male presidents
              avg_accession_age_female_president = mean(
                president_age_at_accession[is_presidential & is_female_president], na.rm = TRUE),
              avg_accession_age_male_president = mean(
                president_age_at_accession[is_presidential & !is_female_president], na.rm = TRUE),
              # calculating difference in average accession ages of male and female presidents (male - female)
              gap_president_accession_age = avg_accession_age_male_president - avg_accession_age_female_president,
              
              # calculating proportions of female and male monarchs
              prop_female_monarch = mean(
                ifelse(is_monarchy, is_female_monarch, NA), na.rm = TRUE),
              prop_male_monarch = mean(
                ifelse(is_monarchy, !is_female_monarch, NA), na.rm = TRUE),
              # calculating difference in proportions of male monarchs and female monarchs (male - female)
              gap_monarch_prop = prop_male_monarch - prop_female_monarch,
              # calculating the average accession ages for female and male monarchs
              avg_accession_age_female_monarch = mean(
                monarch_age_at_accession[is_monarchy & is_female_monarch], na.rm = TRUE),
              avg_accession_age_male_monarch = mean(
                monarch_age_at_accession[is_monarchy & !is_female_monarch], na.rm = TRUE),
              # calculating difference in average accession ages of male and female monarchs (male - female)
              gap_monarch_accession_age =
                avg_accession_age_male_monarch - avg_accession_age_female_monarch,
              .groups = "drop")
}

plotMe <- summarise_leadership_gender_gap("year")
```

```{r}
#| fig-alt: This is a boxplot depicting the proportion of women in parliament, while faceting on the democracy status of the country. The mean proportion of women holding seats in non-democratic countries is 0.16 and 0.177 for democratic countries. The median proportion of women holding seats in non-democratic countries is 0.14 and 0.158 for democratic countries.

pal0 <- wes_palette("Moonrise3", n = 2)  
democracy_data %>%
  filter(!is.na(is_democracy)) %>%
  filter(!is.na(prop_in_par)) %>%
  mutate(is_democracy = factor(is_democracy, labels = c("Non-Democracy", "Democracy"))) %>%
  ggplot(aes(x = is_democracy, y = prop_in_par)) +
  theme(text = element_text(size = 11, family = "serif")) +
  geom_boxplot(fill = pal0,              
    color = "black",          
    outlier.color = "black") +
  stat_summary(fun = mean,
    geom = "crossbar",
    color = "black",
    width = 0.75,
    linetype = "dotted",
    size = .2) +
  stat_summary(fun = mean,
    geom = "text",
    aes(label = paste0("Mean: ", round(after_stat(y), 1),"%")),
    vjust = -1,
    color = "black") +
  stat_summary(fun = median,
    geom = "text",
    aes(label = paste0("Median: ", round(after_stat(y), 1),"%")),
    vjust = 1.5,
    color = "black") +
  labs(title = "Women's Representation in National Parliament by Democracy Status",
    x = "Democracy Status",
    y = "Percentage of National Parliament Seats Held by Women") + 
  theme(plot.title = element_text(size = 5),
    axis.title = element_text(size = 5)) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  theme_minimal()
```

We began by looking at the percentages of women's parliamentary seats in democratic and non-democratic countries. On average, democracies tend to have slightly higher representation, with a median of 15.8% compared to 14.0% in non-democratic countries, and a mean of 17.7% versus 16.0%. That said, the two groups look fairly similar overall, with a lot of overlap between them and wide variation within each group. Non-democratic countries do show a few unusually high values, including two countries where women held more than 60% of parliamentary seats.

To get a closer look at how democracy status relates to gender gaps beyond parliament, we examined how the proportions of male and female presidents and monarchs and the ages at which they take power differ between democratic and non-democratic countries.

```{r}
# Table 1: Proportions
summarise_leadership_gender_gap("is_democracy") |>
  mutate(is_democracy = ifelse(is_democracy, "Democracy", "Non-Democracy")) |>
  select(is_democracy, n,
         prop_female_president, prop_male_president, gap_president_prop, prop_female_monarch, prop_male_monarch, gap_monarch_prop) |>
  gt() |>
  tab_header(
    title    = md("**Gender Gaps in Executive Leadership by Democracy Status**"),
    subtitle = "Proportions of Female and Male Leaders") |>
  tab_spanner(
    label   = "Presidents",
    columns = c(prop_female_president, prop_male_president, gap_president_prop)) |>
  tab_spanner(
    label   = "Monarchs",
    columns = c(prop_female_monarch, prop_male_monarch, gap_monarch_prop)) |>
  cols_label(is_democracy = "Democracy Status",
    n = "Sample Size", prop_female_president = "Prop. Female",
    prop_male_president = "Prop. Male",
    gap_president_prop = "Gap (M - F)",
    prop_female_monarch = "Prop. Female",
    prop_male_monarch = "Prop. Male",
    gap_monarch_prop = "Gap (M - F)") |>
  fmt_percent(columns = c(prop_female_president, prop_male_president, gap_president_prop, prop_female_monarch, prop_male_monarch, gap_monarch_prop),
    decimals = 1) |>
  fmt_integer(columns = n) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_spanners()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_title()) |>
  tab_style(style = cell_fill(color = paste0(pal[1], "99")),
    locations = cells_body(
      columns = gap_president_prop,
      rows = gap_president_prop > 0)) |>
  tab_style(style = cell_fill(color = pal[2]),
    locations = cells_body(
      columns = gap_president_prop,
      rows = gap_president_prop < 0)) |>
  tab_style(style = cell_fill(color = paste0(pal[1], "99")),
    locations = cells_body(
      columns = gap_monarch_prop,
      rows = gap_monarch_prop > 0)) |>
  tab_style(style = cell_fill(color = pal[2]),
    locations = cells_body(
      columns = gap_monarch_prop,
      rows = gap_monarch_prop < 0)) |>
  tab_footnote(footnote  = "Positive values (shaded blue) indicate males are higher; negative values (shaded pink) indicate females are higher.",
    locations = cells_column_labels(columns = c(gap_president_prop, gap_monarch_prop))) |>
  opt_table_font(font = google_font("Source Serif 4"))

# Table 2: Accession Ages
summarise_leadership_gender_gap("is_democracy") |>
  mutate(is_democracy = ifelse(is_democracy, "Democracy", "Non-Democracy")) |>
  select(is_democracy, n,
         avg_accession_age_female_president, avg_accession_age_male_president, gap_president_accession_age,
         avg_accession_age_female_monarch, avg_accession_age_male_monarch, gap_monarch_accession_age) |>
  gt() |>
  tab_header(title = md("**Gender Gaps in Accession Age by Democracy Status**"),
    subtitle = "Average Accession Ages for Female and Male Leaders") |>
  tab_spanner(label = "Presidents",
    columns = c(avg_accession_age_female_president, avg_accession_age_male_president, gap_president_accession_age)) |>
  tab_spanner(label = "Monarchs",
    columns = c(avg_accession_age_female_monarch, avg_accession_age_male_monarch, gap_monarch_accession_age)) |>
  cols_label(
    is_democracy = "Democracy Status",
    n = "Sample Size",
    avg_accession_age_female_president = "Avg. Age (Female)",
    avg_accession_age_male_president = "Avg. Age (Male)",
    gap_president_accession_age = "Gap (M - F)",
    avg_accession_age_female_monarch  = "Avg. Age (Female)",
    avg_accession_age_male_monarch = "Avg. Age (Male)",
    gap_monarch_accession_age = "Gap (M - F)") |>
  fmt_number(columns  = c(avg_accession_age_female_president, avg_accession_age_male_president, gap_president_accession_age, avg_accession_age_female_monarch, avg_accession_age_male_monarch, gap_monarch_accession_age),
    decimals = 1) |>
  fmt_integer(columns = n) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_spanners()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_title()) |>
  tab_style(style = cell_fill(color = paste0(pal[1], "99")),
    locations = cells_body(
      columns = gap_president_accession_age,
      rows = gap_president_accession_age > 0)) |>
  tab_style(style = cell_fill(color = pal[2]),
    locations = cells_body(
      columns = gap_president_accession_age,
      rows  = gap_president_accession_age < 0)) |>
  tab_style(style = cell_fill(color = paste0(pal[1], "99")),
    locations = cells_body(
      columns = gap_monarch_accession_age,
      rows = gap_monarch_accession_age > 0)) |>
  tab_style(style = cell_fill(color = pal[2]),
    locations = cells_body(
      columns = gap_monarch_accession_age,
      rows = gap_monarch_accession_age < 0)) |>
  tab_footnote(footnote  = "Positive values (shaded blue) indicate males are higher; negative values (shaded pink) indicate females are higher.",
    locations = cells_column_labels(columns = c(gap_president_accession_age, gap_monarch_accession_age))) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

Female presidents are more common in democratic countries than in non-democratic ones, while the share of male presidents stays relatively consistent regardless of democracy status. The most notable finding in the accession age table is that female presidents in non-democratic countries take office at nearly 14 years older on average than male presidents. In democratic countries, that difference is much smaller, at under 3 years.

For monarchs, the pattern is quite different. In non-democratic countries, male monarchs outnumber female monarchs by more than 84 percentage points, while in democratic countries the split is much closer to even. Monarchs also tend to come to power younger than presidents, usually somewhere between 30 and 35 years old, and this is especially true in non-democratic countries.

To get a clearer sense of how these age gaps have shifted over time, we constructed a ribbon plot showing the difference between average accession ages for male and female presidents and monarchs by decade. Because there were no recorded female presidents before 1980, we limited this portion of the analysis to years from 1980 onward.

```{r}
#| warning: false
#| fig-alt: This is a ribbon plot, depicting the differences in average accession ages of presidents and monarchs by gender. There are four lines on the graph, one for each gender and leader type combination. The x-axis is years 1980-2020, and the y-axis shows ages 20-70. 

plotMe %>%
  filter(year >= 1980) %>%
  ggplot(aes(x = year)) +
  geom_ribbon(data = plotMe, aes(x = year,
                                 ymin = avg_accession_age_female_president,
                                 ymax = avg_accession_age_male_president),
              fill = "#D3D3D3", alpha = 0.5) +
  geom_ribbon(data = plotMe, mapping = aes(x = year,
                                           ymin = avg_accession_age_female_monarch,
                                           ymax = avg_accession_age_male_monarch),
              fill = "#D3D3D3", alpha = 0.5) +
  geom_line(aes(y = avg_accession_age_female_president, color = "Female President"), linewidth = 1.5) +
  geom_line(aes(y = avg_accession_age_male_president, color = "Male President"), linewidth = 1.5) +
  geom_line(aes(y = avg_accession_age_female_monarch, color = "Female Monarch"), linewidth = 1.5) +
  geom_line(aes(y = avg_accession_age_male_monarch, color = "Male Monarch"), linewidth = 1.5) +
  scale_color_manual(name = "Leader Type",
    values = c("Female President" = pal[3],
      "Male President"= pal[2],
      "Male Monarch" = pal[4],
      "Female Monarch" = pal[1]),
    breaks = c("Female President", "Male President", "Male Monarch", "Female Monarch")) +
  scale_x_continuous(limits = c(1980, 2020)) +
  scale_y_continuous(limits = c(20, 70)) +
  labs(title = "Average Accession Ages of Presidents & Monarchs by Gender", x = "Year", y = "Average Accession Age") +
  theme_minimal()
```

From 1980 to 2020, gender differences in the age at which leaders take office vary between presidents and monarchs. Among presidents, the gap is small and relatively stable, with men only slightly older than women at the time they take office. Both groups also show a gradual increase in age over time. In contrast, monarchs show a much larger and growing gap. Male monarchs tend to take the throne at older ages, while female monarchs are generally younger, with their ages even decreasing in more recent years. Overall, gender disparities in age at taking office are minimal in presidential systems but much more pronounced in monarchical ones.

The next table examines how gender gaps appear in leadership across decades. It compares the proportions of female and male presidents over time to show how representation has shifted from 1950 to 2020.

```{r}
#| results: asis
# table with iteration
group_vars <- c("decade")
group_labels <- c("Decade")

walk2(group_vars, group_labels, function(var, label) {
  tbl <- summarise_leadership_gender_gap(var) |>
    select(all_of(var), n, prop_female_president, prop_male_president, gap_president_prop)
  
  print(tbl |> gt() |>
    tab_header(title = md(paste0("**Gender Gaps by ", label, "**"))) |>
    cols_label(
      !!sym(var)            := label,
      n                      = "Sample Size",
      prop_female_president  = "Prop. Female President",
      prop_male_president    = "Prop. Male President",
      gap_president_prop     = "Gap (M - F)"
    ) |>
    fmt_percent(
      columns  = c(prop_female_president, prop_male_president, gap_president_prop),
      decimals = 1) |>
    fmt_integer(columns = n) |>
    tab_style(style = cell_fill(color = "#D3D3D3"),
      locations = cells_column_labels()) |>
    tab_style(style = cell_fill(color = "#D3D3D3"),
      locations = cells_title()) |>
    opt_table_font(font = google_font("Source Serif 4")))
})
```

This table highlights the changes in gender gaps by decade, including the sample size, decade, proportion of female presidents, and proportion of male presidents. Highlighted in pink are the gender gaps, which as we can predict, decrease in proportion over time. For the first two decades that we have data for (1950-1970), there were solely male presidents, but during our final decade, 2010-2020, there was a gap of around 86%.

We ran a t-test to compare the means of the proportion of national parliament seats held by women between different democracy statuses.

```{r}
result <- t.test(prop_in_par ~ is_democracy, data = democracy_data)
table_result <- tidy(result)

table_result |>
  select(estimate1, estimate2, statistic, p.value, conf.low, conf.high, parameter) |>
  gt() |>
  tab_header(title = md("**Proportion in Parliament by Democracy Status**"), subtitle = "Welch Two-Sample t-Test") |>
  cols_label(estimate1 = md("Mean<br>(Non-Democracy)"),
    estimate2 = md("Mean<br>(Democracy)"),
    statistic = md("*t*-statistic"),
    p.value = md("*p*-value"),
    conf.low = "95% CI Low",
    conf.high = "95% CI High",
    parameter = md("*df*")) |>
  fmt_number(columns  = c(estimate1, estimate2, statistic, conf.low, conf.high),decimals = 3) |>
  fmt_number(columns  = parameter,
    decimals = 1) |>
  fmt_scientific(columns = p.value,
    decimals = 3) |>
  tab_style(style = cell_fill(color = "#FAD77B"),
    locations = cells_body(columns = p.value)) |>
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_body(columns = p.value)) |>
  tab_spanner(label = "Group Means",
    columns = c(estimate1, estimate2)) |>
  tab_spanner(label = "Confidence Interval",
    columns = c(conf.low, conf.high)) |>
  tab_footnote(footnote = "Highlighted cell indicates a statistically significant result.", locations = cells_column_labels(columns = p.value)) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_title()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_spanners()) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

The t-statistic is -4.591 and the p-value is very small. Also, the confidence interval does not include zero, so we reject the null hypothesis and conclude that there is a statistically significant difference in the mean proportion of seats held by women between democratic and non-democratic countries. This suggests that the true difference in means is not zero.

## Exploring The Global Effect of Democratic Transitions

```{r}
#arranging by country and year
democracy_data <- democracy_data |> 
  arrange(country, year)

#creating new columns to the data, and assigning them to 0 to start with. this is an indicator variable, with 0 being no transition, and 1 meaning there is a transition. for example, if for a country 2000 is a 0 for the to_democracy variable and 2001 there is a 1, there was a transition to democracy in 2001. 
democracy_data$to_democracy <- 0
democracy_data$from_democracy <- 0

#to iterate through rows to find out which ones have transitions. starting at 2 because we are comparing to the previous row (2-1 = 1)
for (i in 2:nrow(democracy_data)) {
  #checking if the two countries that are next to each other in the data are the same 
  if (democracy_data$country[i] == democracy_data$country[i-1]) { 
    #assigning them to current and previous labels
    current <- democracy_data$is_democracy[i]
    previous <- democracy_data$is_democracy[i - 1]
    #making sure there are no na values for the is_democracy variable
    if (!is.na(current) && !is.na(previous)) {
      #if this year there is democracy and the previous year there wasn't, it is marked as a transition to democracy
      if (current && !previous) {
        democracy_data$to_democracy[i] <- 1
      }
      #opposite for this one 
    if (!current && previous) {
        democracy_data$from_democracy[i] <- 1
      }
    }
  }
}

#first grouping by year, and then summarizing the proportion of democracy, and transitions in and out of democracy
summary_data <- democracy_data |>
  group_by(year) |>
  summarize(
    prop_democracy = mean(is_democracy, na.rm = TRUE),
    transitions_in = sum(to_democracy, na.rm = TRUE),
    transitions_out = sum(from_democracy, na.rm = TRUE))

#after viewing this data, it would be helpful to analyze these trends using something like a plot, but looking at it as a table is overwhelming. To counteract this, we can condense the years to 5 or 10 years at the time to get a better look at overall trends. 
#we have to create a new column to group the years
democracy_data <- democracy_data |>
  mutate(ten_yrs = floor(year / 10) * 10)
#used the floor function here, that rounds a number down to the greatest integer less than or equal to it, here by 10s.
summary_ten_yrs <- democracy_data |>
  group_by(ten_yrs) |>
  summarize(
    prop_democracy = mean(is_democracy, na.rm = TRUE),
    transitions_in = sum(to_democracy, na.rm = TRUE),
    transitions_out = sum(from_democracy, na.rm = TRUE),
    .groups = "drop")
```

```{r}
ggplot(summary_data, aes(x = year, y = prop_democracy)) +
  geom_point(color = pal[2], size = 2) +
  geom_line(color=pal[2], linewidth=0.75) +
  scale_y_continuous(breaks=seq(0,1,0.2), limits=c(0,1),
                     labels = scales::percent_format()) +
  labs(title = "Percentage of Democratic Countries Over Time",
    x = "Year",
    y = "Percentage of Democratic Countries") +
  scale_x_continuous(breaks=seq(1950,2020,10)) +
  theme_minimal()
```

We first examined the overall trend of the percentage of democratic countries. In 1950, the percentage of democratic countries was at about 25%, before rising gradually through the 1960s. The percentage then stayed fairly stable around the mid-30% range through the 1970s and early 1980s. The largest increase happened around the late 1980s and early 1990s, when the percentage increased sharply to nearly 60%. After that, democracy continued to grow more slowly, reaching the mid-60% range in the 2010s before leveling off slightly by 2020.

Next, we analyzed the number of global transitions to and from democracies.
```{r}
#| results: asis
summary_ten_yrs |>
  select(ten_yrs, transitions_in, transitions_out) |>
  gt() |>
  tab_header(title = md("**Democracy Transitions by Year**"),
    subtitle = "Transitions In and Out of Democracy") |>
  cols_label(ten_yrs = "Decade",
    transitions_in = "Transitions In",
    transitions_out = "Transitions Out") |>
  fmt_integer(columns = c(transitions_in, transitions_out)) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_title()) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

This table shows democratic transitions by decade, offering a cleaner view of long-term trends in global democratization from 1950 to 2020. The 1990s stand out as the most transformative decade, with 39 transitions into democracy and only 11 out, reflecting the wave of democratization that followed the end of the Cold War. In contrast, the 1970s were the most stagnant period, with transitions in and out nearly balanced at 16 and 15 respectively and virtually no net change in the global proportion.

```{r}
#| fig-alt: The area plot below shows the count of democractic transitions in and out, where the pink represents a country gaining a democratic status, and the blue represents a country losing a democratic status throughout the years.

summary_data_long <- summary_data |>
  pivot_longer(
    cols = c(transitions_in, transitions_out),
    names_to = "type",
    values_to = "count")

ggplot(summary_data_long, aes(x = year, y = count, fill = type)) +
  geom_area(alpha = 0.75) +
  scale_fill_manual(values = c(transitions_in = pal[2],
      transitions_out = pal[1]),
    name = "Transition Direction",
    labels = c(transitions_in = "Transitions In",
      transitions_out = "Transitions Out")) +
  labs(title = "Global Democratic Transitions Over Time",
    x = "Year",
    y = "Count of Transitions") +
  scale_y_continuous(limits=c(0,12), breaks=seq(0,12,2)) +
  scale_x_continuous(limits=c(1950,2020), breaks=seq(1950,2020,10)) +
  theme_minimal()
```

The area plot reveals that countries do not transition into democracies at a steady rate, but instead noticeably fluctuate over time. While there are often more transitions into democracy than out of it, this pattern appears variable. The process of democratization is unstable, as transitions out of democracy are present throughout the years and do not seem to be declining. One particularly noticeable feature is seen in the 1990s, where there is a sharp increase in the amount of countries that transitioned into a democracy. We can see that democratization is not a smooth, continuous process. 

We also wanted to identify which countries had the most transitions into and out of democracy from 1950-2020.

```{r}
topn_countries <- function(start_year = 1950, end_year = 2020, num_countries = 10) {
  if (start_year < 1950 | end_year > 2020 | start_year > end_year) {
    stop("Year range must be between 1950 and 2020")
  }
  
  data <- democracy_data |>
    filter(year >= start_year & year <= end_year) |>
    arrange(country_name, year) |>
    group_by(country_name) |>
    mutate(
      prev_year = lag(is_democracy),
      to_democracy = ifelse(is_democracy == 1 & prev_year == 0, 1, 0),
      from_democracy = ifelse(is_democracy == 0 & prev_year == 1, 1, 0)) |>
    ungroup()
  
  country_table <- data |>
    group_by(country_name) |>
    summarize(
      transitions_in = sum(to_democracy, na.rm = TRUE),
      transitions_out = sum(from_democracy, na.rm = TRUE),
      total_transitions = transitions_in + transitions_out,
      .groups = "drop") |>
    arrange(desc(total_transitions)) |>
    slice_head(n = num_countries)
  
  return(country_table)
}

country_transitions <- topn_countries(1950, 2020, 10)

country_transitions |>
  gt() |>
  tab_header(title = md("**Top 10 Countries by Democratic Transitions**"),
    subtitle = "Total Transitions Into and Out of Democracy") |>
  cols_label(country_name = "Country",
    transitions_in = "Transitions In",
    transitions_out = "Transitions Out",
    total_transitions = "Total Transitions") |>
  fmt_integer(columns = c(transitions_in, transitions_out, total_transitions)) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_title()) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

Argentina and Thailand experienced the highest number of democratic transitions between 1950 and 2020, followed by Ghana, Pakistan, and Peru. The countries listed in the table suggest repeated shifts between democratic and non-democratic rule, reflecting political instability over time. Together with the area plot above, this table reinforces that democratization is not a steady or linear process, but one marked by periods of fluctuation.

Lastly, our table earlier showed that the most transitions into democracy happened between 1980-2010. We created another table focusing in on the top 10 countries in this time frame.

```{r}
country_transitions2 <- topn_countries(1980, 2010, 10)
country_transitions2 |>
  gt() |>
  tab_header(title = md("**Top 10 Countries by Democratic Transitions**"),
    subtitle = "Total Transitions Into and Out of Democracy from 1980-2010") |>
  cols_label(country_name = "Country",
    transitions_in = "Transitions In",
    transitions_out = "Transitions Out",
    total_transitions = "Total Transitions") |>
  fmt_integer(columns = c(transitions_in, transitions_out, total_transitions)) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_title()) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

In this table and the previous one, we see that the countries listed all have four or fewer transitions in or out of democracy in the respective time frames. Although this is not a very large number, even a small number of regime changes can indicate meaningful political instability, since transitions between democratic and non-democratic systems often reflect major shifts in leadership and ideology. This suggests that a key part of democratization is not only becoming democratic, but also maintaining democratic rule over time. 

## Conclusion

Overall, our analysis shows that democracy is related to gender representation and political stability, but not always in simple ways. Democratic countries had slightly higher average proportions of women in parliament than non-democratic countries, and the t-test showed that this difference was statistically significant. However, the actual difference in means was fairly small, suggesting that democracy alone does not guarantee equal representation. Female presidents became more common over time, but male presidents still made up the large majority of presidential leaders across every decade. Accession-age gaps were smaller and more stable among presidents, while monarchs showed larger differences by gender.

Our analysis of democratic transitions also showed that global democratization has been unpredictable. The percentage of democratic countries increased substantially from 1950 to 2020, especially around the late 1980s and early 1990s, but this growth was not steady. Countries continued to move both into and out of democracy over time, showing that democratization includes both progress and reversal. The country-level transition tables suggest that even a small number of regime changes can reflect meaningful political instability.

Taken together, these findings show that becoming a democratic country can be associated with greater gender representation and broader political change, but maintaining democratic rule and achieving gender equality remain ongoing challenges. Democracy may create more opportunities for representation, but the persistence of gender gaps and repeated regime transitions show that political progress is not automatic or permanent.

